---
title: "wrangle"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages
```{r echo=FALSE}
### Load data from Our World in Data
library(dplyr)
library(zoo)
library(ggplot2)
library(openxlsx)

theme_set( theme_bw() + theme( legend.position="bottom" ) +
             theme( legend.title=element_blank() ) +
             theme(plot.title = element_text(hjust = 0.5) ) +
             theme( text = element_text(size=12) ) )

# Assumptions
# rate of doses lost
waste_rate <- 0.000
# average time between 1st and 2nd dose
t1st2nd <- (56+84)/2 #8 to 12 weeks for Oxford AstraZenca

# Constants
dfmts <- c('%Y-%m-%d', '%m/%d/%Y')
Ecountries <- c('Ukraine','Slovenia')
nper <- 14 # days to smooth supply

# truth URL from the CDC ensemble 
vurl <- 'https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/vaccinations/vaccinations.csv'
df <- read.csv( vurl )
save(df, file = "data/owid.rda")
```

Calculate daily history
```{r echo=FALSE}
# (current capacity to administer doses per day is in daily_vaccinations - 7-day moving avg)
dcap <- df %>% filter( location %in% Ecountries ) %>% 
  group_by( location ) %>%
  mutate( date = as.Date(date, tryFormats = dfmts)
          , total_vaccinations = na.approx( total_vaccinations, na.rm=FALSE ) #replace NA by interpolation
          , people_vaccinated_per_hundred = na.approx( people_vaccinated_per_hundred, na.rm=FALSE )
  ) %>% ungroup()
latest_history_date <- max(dcap$date)
print( paste('Latest data from', latest_history_date) )
```

Totals
```{r echo=FALSE}
tmp <- lapply( Ecountries, function( state ){
  idx <- dcap$location==state & dcap$date==latest_history_date
  print( paste('Total Administered', state, dcap$total_vaccinations[idx]) )
  print( paste('People with 1+ Doses', state, dcap$people_vaccinated[idx]) )
  print( paste('People with 2+ Doses', state, dcap$people_fully_vaccinated[idx]) )
  print( paste('Total Administered per 100k', state, dcap$total_vaccinations_per_hundred[idx]*1000) )
  print( paste('People with 1+ Doses per 100k', state, dcap$people_vaccinated_per_hundred[idx]*1000) )
  print( paste('People with 2+ Doses per 100k', state, dcap$people_fully_vaccinated_per_hundred[idx]*1000) )
} )
```

CDC population numbers
```{r echo=FALSE}
pop <- dcap %>% mutate( pop1 = total_vaccinations/(total_vaccinations_per_hundred/100)) %>%
      group_by( location ) %>% summarise( m=mean(pop1, na.rm=TRUE), s1=sd(pop1, na.rm=TRUE), .groups='drop_last' )
print(pop)
```

Ukraine population 44.39M
```{r echo=FALSE}
pop$m[which(pop$location =="Ukraine")] <- 44390000
```

Plot daily_vaccinations (it is already 7-day smoothed in the original data)
```{r echo=FALSE}
dp <- dcap %>% filter( date >= "2021-03-01" ) %>% select(date,location,daily_vaccinations) 
ggplot( dp, aes(x = date, y=daily_vaccinations) ) +
  facet_wrap(~ location, nrow=2, scale='free_y' ) +
  geom_line() + 
  theme(axis.title.x = element_blank()) +
  scale_x_date(date_labels = '%b%e' ) +
  ylab('Daily Vaccinations') +
  ggtitle('EUCOM') +
  scale_y_continuous(label = scales::unit_format(unit = "K", scale = 1e-3, sep = "")
                     , expand = c(0, 1000), limits = c(1000, NA))
```

Forecasting: 
```{r echo=FALSE}
# Forecast function
fcastfn <- function( l, c, al, be ){
  #setup
    fdays <- as.Date('2021-10-15') - latest_history_date # number of days to forecast
    dt <- dcap %>% filter( location == l ) %>% 
      select( all_of(c('date',c)) ) 
    z <- read.zoo( dt )
    zz <- z
    time(zz) <- seq_along(time(zz))
  # forecast model
    sm <- HoltWinters( as.ts(zz[-(1:(nper-1))]), alpha=al, beta=be, gamma=FALSE )
    print( paste('location', l, 'alpha', sm$alpha, 'beta', sm$beta, 'level', sm$coefficients['a'], 'trend', sm$coefficients['b']) )
    pr <- predict(sm, fdays, prediction.interval=TRUE, level = 0.95)
    dates <- seq(max(dt$date)+1, by = "day", length.out = fdays)
    prdf <- data.frame( dates, pr ) 
    prdf[prdf < 0] <- 0    # override negative predictions
    cols <- paste0(c, '_', colnames(prdf[1,-1]))
    colnames(prdf) <- c('date',cols)
    res <- prdf %>% mutate( 'location' = l )
}
print( paste('Latest data from', latest_history_date) )
```

Forecasts
```{r echo=FALSE}
f <- list( )
beta_vaccinations <- 0.03 # Beta forced because of recent spikes in NH
```

Forecast Cumulative number of doses **administered**
```{r echo=FALSE}
print( paste('Latest data from', latest_history_date) )
tmp <- lapply( Ecountries, fcastfn, 'total_vaccinations', NULL, beta_vaccinations )  # first argument, function, optional arguments
f[[1]] <- bind_rows( tmp ) 
```

Forecast **Daily** number of doses **administered**
```{r echo=FALSE}
print( paste('Latest data from', latest_history_date) )
# Forecast daily vaccinations
tmp <- lapply( Ecountries, fcastfn, 'daily_vaccinations', NULL, beta_vaccinations ) # first argument, function, optional arguments
f[[2]] <- bind_rows( tmp ) %>% select(-date,-location)
```

Forecast people vaccinated per capita
```{r echo=FALSE}
print( paste('Latest data from', latest_history_date) )
tmp <- lapply( Ecountries, fcastfn, 'people_vaccinated_per_hundred', NULL, beta_vaccinations ) # first argument, function, optional arguments
f[[3]] <- bind_rows( tmp ) %>% select(-date,-location)
```

Bind history and prediction
```{r echo=FALSE}
fcast <- bind_cols( f ) # JCwhat structure to what structure?
latest_fcast_date <- max(fcast$date)
dcap_s <- dcap %>% select( date, location, daily_vaccinations, total_vaccinations, people_vaccinated_per_hundred)
fcast_s <- fcast %>% select(date, location, daily_vaccinations_fit, total_vaccinations_fit, people_vaccinated_per_hundred_fit)
colnames(fcast_s) <- colnames(dcap_s)
dfw <- bind_rows( dcap_s, fcast_s ) 
  #mutate( daily_available = total_distributed*(1-waste_rate) - total_vaccinations ) # total availability level
```

Plot Forecast of Cumulative number of doses distributed & administered
```{r echo=FALSE}
colors <- c('daily_vaccinations','total_vaccinations')
dp <- dfw %>% filter( date <= (latest_history_date+14) ) %>%
  tidyr::pivot_longer( cols = all_of(colors), names_to="s", values_to="value", values_drop_na=TRUE ) #change "colors" vars into 1 column with associated value
cnames <- c('Daily','Cumulative')
dp <- dp %>% mutate( n = NA ) #add new variable n
dp$n[dp$s %in% colors] <- cnames[match(dp$s,colors)] #assign a cname to each 
ggplot( dp, aes(x = date, y=value, color=n) ) +
  facet_wrap('~ location', nrow=2, scale='free_y' ) +
  geom_line() + 
  geom_vline(xintercept=latest_history_date, linetype="dashed", 
             color = "gray") +  
  ggtitle( paste('Latest data from', latest_history_date) ) +
  theme(axis.title.x = element_blank()) +
  scale_x_date( date_labels = '%b%e' ) +
  ylab('Cummulative Number of Doses') +
  scale_y_continuous(label = scales::unit_format(unit = "K", scale = 1e-3, sep = "")
                     , expand = c(0, 0), limits = c(0, NA))
```

Plot Forecasted Cumulative number of people vaccinated per hundred
```{r echo=FALSE}
ggplot( dp, aes(x = date, y=people_vaccinated_per_hundred) ) +
  facet_wrap('~ location', nrow=2 ) +
  geom_line() + 
  geom_vline(xintercept=latest_history_date, linetype="dashed", 
             color = "gray") +  
  ggtitle( paste('Latest data from', latest_history_date) ) +
  theme(axis.title.x = element_blank()) +
  scale_x_date(date_labels = '%b%e' ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA))
```



# CALCULATIONS for Shiny App
```{r}
# capacity assumptions
sc <- left_join(sites, sitecap, by=c('date','location','type')) 

# main dataframes
colors <- c('daily_vaccinations','cap')
cnames <- c('Doses administered','Max vaccination capacity')
dp <- dfw %>% filter( date <= (latest_history_date+14) ) %>% 
    tidyr::pivot_longer( cols = all_of(colors), names_to="s" 
                         , values_to = "value",  values_drop_na = TRUE )
dp <- dp %>% mutate( n = NA )
dp$n[dp$s %in% colors] <- cnames[match(dp$s,colors)]

# per capita values per state and Region 1 average
dpp <- data.frame()
for( var in c('total_vaccinations_per_hundred','people_vaccinated_per_hundred') ){
  tmp <- dfw %>% select( c('date','location',all_of(var)) ) %>%
                       tidyr::pivot_wider( names_from='location', values_from=var ) %>%
                       mutate( avgr1 = (.[[2]]+.[[3]]+.[[4]]+ .[[5]]+ .[[6]]+ .[[7]]) /6 ) %>%
                       tidyr::pivot_longer( cols=all_of(NEstates), names_to="location" 
                                            , values_to = "value",  values_drop_na=FALSE ) %>%
                       tidyr::pivot_longer( cols=c(value,avgr1), names_to="type" 
                                            , values_to=var, values_drop_na=FALSE ) 
  if ( nrow(dpp) == 0 )
    dpp <- tmp
  else
    dpp[,var] <- tmp[,var]
}
dpc <- dpp %>% filter( date <= (latest_history_date+14) )

# Perc people vaccinated over 16 yr old
perc_pop_over16 <- data.frame( state = NEstates,
                               perc_over16 = c(.821, .837, .827, .836, .83, .84) )
for ( st in NEstates ){
  idx <- dpp$location == st
  dpp$people_vaccinated_per_hundredover16[idx] <- dpp$people_vaccinated_per_hundred[idx] / perc_pop_over16$perc_over16[perc_pop_over16$state==st]
}
  
# dates where given vars will exceed given percentages of population
calc_threshold_dates <- function(st, vartype){
  tmp <- lapply( c('people_vaccinated_per_hundred','people_vaccinated_per_hundredover16'), function(colref){
    thrs <- c(70, 90)
    dates <- sapply( thrs, function(threshold){
      datetmp <- dpp %>% filter( type==vartype & location==st & !!as.symbol(colref)>=threshold )
      stopifnot( nrow(datetmp) > 0 )
      return( min( datetmp$date ) )
    } )
    return( data.frame( location=st, type=vartype, colref, threshold=thrs, date=as.Date(dates) ) )
  } ) %>% bind_rows()
}
datesdf <- lapply( NEstates, calc_threshold_dates, 'value' ) %>% bind_rows() %>% tidyr::pivot_wider( names_from=colref, values_from=date )
datesdfavg <- calc_threshold_dates( 'Connecticut', 'avgr1' )

# forecasted growth in daily vaccinations
dvacgr <- dfw %>% group_by( location ) %>%
                summarise( b = last(dd_vaccinations), .groups='drop_last' )
```
```{r}
# test only
tmp <- dpp %>% filter( type=='value' & location==NEstates[4]  )
tmp
```


## SAVE data for Shiny App app.R and for in-transit graph of pre_proc_Tiberius.Rmd
```{r echo=FALSE}
print( paste('SAVING Latest data from', latest_history_date) )
save(dfw, dp, dpp, dpc, datesdf, datesdfavg, dvacgr, latest_history_date, sc, pop, NEstates, stock7d,  file='app/app.RData')
```

